---
title: "Portland Census and ACS Survey Analysis"
author: "Aung Myo Htut"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: architect
    #code_folding: show
    toc: yes
    # toc_float:
    #   collapsed: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

```


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(tidycensus)
library(tigris)
library(sf)
library(crsuggest)
library(mapview)
library(purrr)
library(leaflet.extras2)
library(leafsync)
library(tmap)
library(spdep)
library(viridisLite)
library(htmlwidgets)
library(RColorBrewer)
options(tigris_use_cache = TRUE)
``` 

# INTRODUCTION
The decennial census and the American Community Survey (ACS) are important tools used by the U.S. Census Bureau for demographic data collection, each serving distinct purposes. The decennial census, conducted every ten years, aims to count every person residing in the United States and gather fundamental demographic information such as age, sex, and race. This comprehensive count is essential for apportioning congressional representation and allocating federal funds. In contrast, the ACS is an annual survey that collects detailed demographic, social, economic, and housing data from a sample of the population. While the decennial census provides a complete enumeration, the ACS offers deeper insights into aspects like educational attainment and housing costs, supporting informed policy development, program administration, and business planning. I utilized both the decennial census and ACS data to map various characteristics for the city of Portland, Maine, enhancing our understanding of its demographic landscape and informing local decision-making.


# DECENNIAL CENSUS


```{r Cumber County Shape file Tracts}
# Load Cumberland County Shape File
cumberland_tracts <- tracts(state = "ME", county = "Cumberland", cb = TRUE, year = 2022)
```


```{r Load Variables}
# Census Variables
vars_dhc <- load_variables(2020,"dhc")
vars_dp <- load_variables(2020, "dp")
vars_ddhcca <- load_variables(2020, "ddhca")

# ACS Variables
vars_acs1 <- load_variables(2022,"acs1")
vars_acs5 <- load_variables(2022, "acs5")
vars_profile <- load_variables(2022,"acs5/profile") #data profile for variable lookup
vars_subject <- load_variables(2022, "acs5/subject") #subject tables
vars_cprofile <- load_variables(2022, "acs5/cprofile") #comparison profile
```

## MEDIAN AGE BY BOTH SEXES
```{r median age}
# Cumberland Tracts for Median Age by Both Sex
cumberland_medianage <- get_decennial(
  geography = "tract",
  variables = "DP1_0073C",
  state = "ME",
  county = "Cumberland",
  geometry = TRUE,
  year = 2020,
  sumfile = "dp"
) 

# County Subdivision for Race population percent
subdiv_medianage <- get_decennial(
  geography = "county subdivision",
  variables = "DP1_0073C",
  state = "ME",
  county = "Cumberland",
  geometry = TRUE,
  year = 2020,
  sumfile = "dp") %>% 
  filter(str_detect(NAME, "^Portland"))

# Ensure CRS matches for accurate spatial operation
portland_transform <- st_transform(subdiv_medianage, st_crs(cumberland_medianage))

# Use st_intersection to find tracts within Portland
portland_interset <- st_intersection(cumberland_medianage, portland_transform)

# Erase Water Data
portland_medianage_erase <- erase_water(
  portland_interset,
  year = 2020
)

# Mapping
mapview(portland_medianage_erase,
        zcol = "value",
        layer.name = "Median Age by Both Sexes in <br/> Portland City")
```


```{r Portland Shapefile by Tracts, include=FALSE}
## Creating Portland Shape File
portland_select <- portland_interset %>% 
  select(c("GEOID", "NAME", "NAME.1","geometry"))

# Filter tracts
portland_tracts <- cumberland_tracts[cumberland_tracts$GEOID %in%  portland_select$GEOID,]


#Export to Shape File
#st_write(portland_tracts, "portland.shp")
```


## MALE AND FEMALE POPULATION 
```{r male and female pop}
# Cumberland Tracts for Male and Female Population
cumberland_mfpop <- get_decennial(
  geography = "tract",
  variables = c(Male = "DP1_0025C", Female = "DP1_0049C"),
  state = "ME",
  county = "Cumberland",
  geometry = TRUE,
  year = 2020,
  sumfile = "dp",
  output = "wide"
) 

# Calculate the percentage of males and females
cumberland_mfpop <- cumberland_mfpop %>%
  mutate(
    Total = Male + Female, 
    Percent_Male = round((Male / Total) * 100,2),
    Percent_Female = round((Female / Total) * 100,2)
  )

#Filter to keep only the rows that have matching GEOIDs
portland_mfpop <- cumberland_mfpop %>% 
  filter(GEOID %in% portland_tracts$GEOID)
  
portland_mfpop_erase <- erase_water(
  portland_mfpop,
  year = 2020
)

m <- mapview(portland_mfpop_erase, 
             zcol = "Percent_Male",
             at = c(30,50,70),
             layer.name = "% Male Population",
             col.regions = c("goldenrod","green"))
f <-mapview(portland_mfpop_erase,
            zcol = "Percent_Female", 
            layer.name = "% Female Population",
            at = c(30,50,80),
            col.regions = c("blue","skyblue"))

m | f

```


## TOTAL POPULATION
```{r total pop}
# First Total Population in Cumberland County
cumb_pop <- get_decennial(
  geography = "tract",
  variables = "P1_001N",
  year = 2020,
  state = "ME",
  county = "Cumberland",
  sumfile = "dhc",
  geometry = TRUE
) %>% 
  arrange(desc(value))


# Filter GEOIDs that match in portland_tracts
portland_pop <- cumb_pop %>% 
  filter(GEOID %in% portland_tracts$GEOID)


#Erase Water Area
portland_pop_erase <- erase_water(portland_pop,year = 2020)

colors <- plasma(n= 4)
# Produce Map
mapview(portland_pop_erase, zcol = "value",
        at = seq(1000,9000,2000),
        legend = TRUE,
        col.regions = colors,
        layer.name = "Portland Population")

```


# AMERICAN COMMUNITY SURVEY

## MEDIAN HOUSEHOLD INCOME
```{r income}
cumberland_income <- get_acs(
  geography = "tract",
  variables = "DP03_0062", # Estimate total households with food stamp/snap benefits in the past 12 months
  year = 2022,
  state = "ME",
  county = "Cumberland",
  geometry = TRUE
)

# Filter GEOIDs that match in portland_tracts
portland_income <- cumberland_income %>% 
  filter(GEOID %in% portland_tracts$GEOID)

# mean(portland_income$estimate)
# min(portland_income$estimate)
# max(portland_income$estimate)

#Erase water area
portland_income_erase <- erase_water(
  portland_income,
  area_threshold = 0.5,
  year = 2020
)

#Mapping
c1 = plasma(n=5, direction = -1)

# ggplot(portland_income_erase) +
#   geom_sf(color = "white", aes(fill = estimate)) +
#   geom_sf_text(aes(label = estimate), size = 2.5, color = "yellow")+
#   theme_void() 

mapview(portland_income_erase,
        zcol = "estimate",
        at = seq(10000,110000,20000),
        col.regions = c1,
        layer.name = "Median Household Income <br/> in Portland")
```

## HOUSEHOLDS WITH SNAP BENEFITS
```{r snap}
cumberland_snap <- get_acs(
  geography = "tract",
  variables = "DP03_0074P",
  state = "23",
  county = "005",
  year = 2022,
  geometry = TRUE
)

# Filter GEOIDs that match in portland_tracts
portland_snap <- cumberland_snap %>% 
  filter(GEOID %in% portland_tracts$GEOID)

#Erase Water Area
portland_snap_erase <- erase_water(
  portland_snap,
  year = 2020
)

#Mapping
c2 = rocket(n=6, alpha = 1, direction = -1)
mapview(portland_snap_erase,
        zcol = "estimate",
        at = seq(0,30,5),
        col.regions = c2,
        layer.name = "% of Households with SNAP Benefits <br/> in Portland")
```


## HEALTH INSURANCE COVERAGE
```{r insurance coverage}
Cumberland_insurance <- get_acs(
  geography = "tract",
  variables = "DP03_0096P",
  state = "23",
  county = "005",
  year = 2022,
  geometry = TRUE
)

# Filter GEOIDs that match in portland_tracts
portland_insurance <- Cumberland_insurance %>% 
  filter(GEOID %in% portland_tracts$GEOID)

#Erase Water Area
portland_insurance_erase <- erase_water(
  portland_insurance,
  year = 2020
)

#Mapping
c2 = viridis(n=6, alpha = 1, direction = -1)
mapview(portland_insurance_erase,
        zcol = "estimate",
        at = seq(88,100,2),
        col.regions = c2,
        layer.name = "% of Households with Health Insurance <br/> Coverage in Portland")

```

## POPULATION OVER 25 YRS WITH BACHELOR'S DEGREE

```{r education}
Cumberland_education <- get_acs(
  geography = "tract",
  variables = "DP02_0065P",
  state = "23",
  county = "005",
  year = 2022,
  geometry = TRUE
)

# Filter GEOIDs that match in portland_tracts
portland_education <- Cumberland_education %>% 
  filter(GEOID %in% portland_tracts$GEOID)

#Erase Water Area
portland_education_erase <- erase_water(
  portland_education,
  year = 2020
)

#Mapping
c3 = mako(n=5, alpha = 1, direction = -1)
mapview(portland_education_erase,
        zcol = "estimate",
        at = seq(25,50,5),
        col.regions = c3,
        layer.name = "% of Population with Bachelor's <br/> Degree in Portland")

```

## DISABILITY STATUS
```{r disability}
# Disability status of Civilian Noninstitutionalized Population

Cumberland_disability <- get_acs(
  geography = "tract",
  variables = "DP02_0076P",
  state = "23",
  county = "005",
  year = 2022,
  geometry = TRUE
)

# Filter GEOIDs that match in portland_tracts
portland_disability <- Cumberland_disability %>% 
  filter(GEOID %in% portland_tracts$GEOID)

#Erase Water Area
portland_disability_erase <- erase_water(
  portland_disability,
  year = 2020
)

#Mapping
c4 = magma(n=5, alpha = 1, direction = -1)
mapview(portland_disability_erase,
        zcol = "estimate",
        at = seq(0,30,5),
        col.regions = c4,
        layer.name = "% of Population 18-64 years <br/> with Disabiltiy in Portland")
```

## LANGUAGE SPOKEN
```{r language}
# % of Population 5 years and over language other than English
Cumberland_language <- get_acs(
  geography = "tract",
  variables = "DP02_0114P",
  state = "23",
  county = "005",
  year = 2022,
  geometry = TRUE
)

# Filter GEOIDs that match in portland_tracts
portland_language <- Cumberland_language %>% 
  filter(GEOID %in% portland_tracts$GEOID)

#Erase Water Area
portland_language_erase <- erase_water(
  portland_language,
  year = 2020
)

#Mapping
c5 = plasma(n=6, alpha = 1, direction = -1)
mapview(portland_language_erase,
        zcol = "estimate",
        at = seq(0,30,5),
        col.regions = c5,
        layer.name = "% of Population 5 years and over with language <br/> other than English in Portland")
```

## FAMILY INCOME BELOW POVERTY LEVEL
```{r poverty}
# % of families and people whose income below the poverty level in the past 12 months
Cumberland_poverty <- get_acs(
  geography = "tract",
  variables = "DP03_0119P",
  state = "23",
  county = "005",
  year = 2022,
  geometry = TRUE
)

# Filter GEOIDs that match in portland_tracts
portland_poverty <- Cumberland_poverty %>% 
  filter(GEOID %in% portland_tracts$GEOID)

#Erase Water Area
portland_poverty_erase <- erase_water(
  portland_poverty,
  year = 2020
)

#Mapping
#c5 = plasma(n=6, alpha = 1, direction = -1)
mapview(portland_poverty_erase,
        zcol = "estimate",
        at = seq(0,15,3),
        col.regions = brewer.pal(5,"YlOrBr"),
        layer.name = "% of Families And People Below <br/> Poverty Level in Portland")
```

## RENTER OCCUPIED HOUSING
```{r housing}
# Renter Occupied Housing Units
Cumberland_rent <- get_acs(
  geography = "tract",
  variables = "DP04_0047P",
  state = "23",
  county = "005",
  year = 2022,
  geometry = TRUE
)

# Filter GEOIDs that match in portland_tracts
portland_rent <- Cumberland_rent %>% 
  filter(GEOID %in% portland_tracts$GEOID)

#Erase Water Area
portland_rent_erase <- erase_water(
  portland_rent,
  year = 2020
)

#Mapping
#c5 = plasma(n=6, alpha = 1, direction = -1)
mapview(portland_rent_erase,
        zcol = "estimate",
        at = seq(0,100,20),
        col.regions = brewer.pal(5,"Blues"),
        layer.name = "% of Renter Occupied Housing <br/> Units in Portland")
```

